% !TeX spellcheck = en_GB

% =================================================================
\chapter{Introduction}
\label{chap:intro}

\begin{itemize}
  \item
    see Exposé
  \item
    Example queries:
    %
    \begin{enumerate}
      \item[\exaquery{1}]
        Who read work $W$, in which manifestation and in which year?
      \item[\exaquery{2}]
        Which copies of work $W$ were passed from one of its owners to a collaborator (or a student)?
      \item[\exaquery{3}]
        What are the relationships between the recipients of work $W$
        (or of manifestation $M$ of $W$ or of copy $C$ of $W$, respectively)?
    \end{enumerate}
\end{itemize}
%
To illustrate these abstract queries and possible approaches to answering them, let us fix work $W$ to be the seminal work \emph{De revolutionibus orbium coelestium}
(short: \emph{De revolutionibus}; English translation: \emph{On the Revolutions of the Heavenly Spheres}) by the astronomer Nicolaus Copernicus (1473–1543) \todo{citation}.
We consider the following more precise variant of \exaquery{2}:
%
\begin{enumerate}
  \item[\exaquery{2$'$}]
    Which copies of \emph{De revolutionibus} were owned by scientists who passed them on to a student?
\end{enumerate}
%
Let us first assume that a researcher wants to answer this question.
One obvious way to proceed would be as follows: first, our researcher finds copies of \emph{De revolutionibus} 
in online catalogues of libraries and library networks. For each such copy, they then inspect the provenance entries
that name owners who were people (not corporations). Finally, our researcher will have to find those names in databases such as
authority files or Wikidata and, for each entry, explore the specified profession (``scientist'') and relationships to other people (``student'').

For example, the online catalogue (OPAC) of the Gotha Research Library of the University of Erfurt (Forschungsbibliothek Gotha) lists two printed copies
of \emph{De revolutionibus}.%
\footnote{%
  \url{https://opac.uni-erfurt.de/DB=1/CMD?ACT=SRCHA&IKT=1016&SRT=YOP&TRM=tit+de+revolutionibus+and+per+kopernikus+and+jah+15**+and+bbg+a*}%
}
One of these copies bears the signature \sig{Druck~4°~00466}, and its provenance entries name the following previous owners:%
\footnote{%
  \url{https://opac.uni-erfurt.de/DB=1/XMLPRS=N/PPN?PPN=567506266}%
}
%
\begin{itemize}
  \item
    Hieronymus Tilesius (1529–1566): autograph and date 1551
  \item
    NN: note, date 1553, name scraped out
  \item
    Johann Hommel (1518–1562), autograph
  \item
    Valentin Thau (1531–1575), note (greek proverb, possibly not denoting ownership)
  \item
    Ernest II, Duke of Saxe-Gotha-Altenburg (1745–1804): stamp/seal, initial
  \item
    Ducal Library, Gotha (a predecessor organisation of Gotha Research Library): stamp marking a duplicate
  \item
    Ernestine Gymnasium, Gotha: stamp
  \item
    Landesbibliothek Gotha: stamp
\end{itemize}
%
Our researcher can immediately decide that they can ignore the second entry (no name given) and the last three entries (corporations).
For the remaining four entries, our researcher follows the links given in the OPAC to the Integrated Authority File (GND) of the German National Library.%
\footnote{%
  \url{https://katalog.dnb.de/EN/home.html?v=plist}%
}
On inspection of these entries, it turns out that Ernest II was a regent and very probably not a scientist,
and that the other three people---Tilesius, Hommel, and Thau---had professions such as theologian,
mathematician, and astronomer, which qualifies them as scientists. Furthermore, Hommel's entry
contains a reference to Thau with the relationship ``has student''
(and Thau's entry contains the inverse reference to Hommel).
From this reference, our researcher can conclude that two scientists in the teacher-student relation
have both possessed the book. Since the data available does implies neither that Hommel passed the book (directly) to Thau
nor that Thau was really an owner,
our researcher can formulate hypotheses based on the retrieved data and start an in-depth research.

The manual process that we have just described is cumbersome, laborious, and prone to errors and omissions for several reasons:
The search in library catalogues for copies of works and their provenances requires expert skills.
Catalogues with potential matches need to be selected manually,
and each catalogue needs to be queried individually, using its own search functionality and syntax. \todo{give examples of commonalities (OPAC) and differences (discovery vs. OPAC)?}
The traversal through all retrieved copies and the pursuit of each potential relevant provenance entry per copy 
multiplies the amount of manual work necessary.
Finally, it is not clear what an effective and efficient way to ``explore'' relationships would be:
while it is easy to find direct relationships such as ``person $P_1$ is student of person $P_2$'' in the view for a person's entry
in databases such as GND or Wikidata, there are relationships that cannot be discovered easily by hand,
e.g., ``$P_1$ and $P_2$ are students of the same scholar''.

It should be clear from this example that the process of answering questions such as \exaquery{2$'$}
would strongly benefit from automated support, which could help reduce the amount of manual work, integrate heterogeneous data sources,
incorporate background knowledge (e.g., every mathematician is a scientist),
and discover relationships between entities that are not necessarily direct.
It is also clear that, due to the incompleteness of data, query answers obtained by manual or automated search
are always tentative in two complementary senses: First, if relevant provenance entries or relationships between people are missing
in the data sources, then neither a human nor a machine will be able to retrieve the respective answers. Hence, the sets of query answers obtained
are not necessarily complete.
Second, since the relationship ``person $P_1$ passed a book on to another person $P_2$'' is generally omitted from
provenance entries, a human or machine can only retrieve possible constellations such as the ones
given above. Hence, not all query answers need to be correct, but all answers can be considered
as informed hypotheses that initiate further research.

\todo[inline]{!! ``scientist'' isn't given in GND for, e.g., Thau -- only more specific relations such as ``mathematician'', ``astronomer'', ``lawyer''. $\Rightarrow$ we need QA with taxonomies/ontologies !!}

...
%
\begin{itemize}
  \item
    more examples for \exaquery{1} or \exaquery{3}?
  \item
    explain how \exaquery{1}--\exaquery{3} differ from each other
  \item
    In this thesis, focus on these three types of queries.
    This is a deliberate restriction of what one can generally ask – scope of master's thesis!
\end{itemize}



\dots

